Concurrent Programming with CXL

### 2020-05-18 book/paper.tex
Latest
Working on multi-argument methods

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
possible inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 10, different values, then one can hope that it
will also work for sets that are larger than 10 values.

@@

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point or lists natively;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The $x..y$ notation generates a set containing the numbers
from $x$ to $y$ (inclusive).  The last two lines in the program are
the most interesting.
They assign to $x$ some unspecified value in the range $0..N$, and then
verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.
If the verification fails, the value of $x$ is reported.
(The syntax of a CXL \texttt{assert} statement is the same as in Python
except that the ``,'' clause is not optional.)

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Essentially, the $\texttt{choose}(S)$
operator provides the input to the program by selecting some value from the
set $S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' and its diameter (more on this later).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
to \texttt{False}.
The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the next chapter, but suffice it to
say that The machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes 0 and 1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

The result in this particular interleaving is that \texttt{count} ends up
being 1.  When running CXL, it will
report the violations of the assertions.  It also provides an example
of an interleaving, like the one above, of how exactly the assertion
ended up failing.

@@

\begin{code}
\begin{verbatim}
    assert (count == 0) or (count == 1), count;
\end{verbatim}
\end{code}

@@

A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable \emph{name tag} (specified when spawning the
process), a program counter, a stack of CXL values, and a single mutable
general purpose \emph{register} that contains a CXL.
\end{itemize}

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

>>>>

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
possible inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 10, different values, then one can hope that it
will also work for sets that are larger than 10 values or sets that
use more than 10 values.

@@

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The $x..y$ notation generates a set containing the numbers
from~$x$ to~$y$ (inclusive).  The last two lines in the program are
the most interesting.
They assign to $x$ some unspecified value in the range $0..N$, and then
verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.
If the verification fails, the value of $x$ is reported.

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Essentially, the $\texttt{choose}(S)$
operator provides the input to the program by selecting some value from the
set~$S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' and its diameter (more on this later).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the next chapter, but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
The following is a possible interleaving of processes~0 and~1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

The result in this particular interleaving is that \texttt{count} ends up
being 1.  When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

\begin{code}
\begin{verbatim}
    assert (count == 0) or (count == 1);
\end{verbatim}
\end{code}

@@

A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{dict\{ .name: .\_\_main\_\_, .tag: () \}}, pc 0,
register \texttt{()}, and an empty stack.
(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

\chapter{Using CXL Methods}
